# The `causl` Package {#causl}

In this chapter we introduce the `causl` package, which performs basic simulation
and fitting for static models, as discussed in @evans23param.


## Installation

You can obtain and install `causl` from GitHub.  You will need the `devtools` 
package 
```{r causl_install, eval=FALSE}
install.packages("devtools")
devtools::install_github("rje42/causl") 
## or remotes::install_github("rje42/causl")
```

Now load the library.
```{r causl_load}
library(causl)
```

You can immediately simulate from your first model:
```{r basic_causl}
rfrugalParam(10, pars=list(z=list(beta=0, phi=1),
                           x=list(beta=c(0,0.5), phi=0.75),
                           y=list(beta=c(-0.25,0.5), phi=0.5),
                           cop=list(beta=1)),
             family=c(1,1,1,1))
```

## Syntax

Specify a collection of formulas to define your model.
```{r formulas_causl}
forms <- list(Z ~ A,
              list(A ~ 1, B ~ A*Z),
              Y ~ A*B,
               ~ 1)
```
Note that multiple formulas can be specified by putting a list into the relevant
entry.

You will also need to specify a list of the family variables.
```{r family_causl}
fam <- list(1, c(5,5), 1, 1)
```
Note that we can have a vector of family variables if one of the entries (covariates,
treatments, outcomes, copula) has multiple entries.  Here 1 indicates a Gaussian
distribution, and 5 a Bernoulli.  

We also need to specify parameters for each random variable, and for the copula.
```{r pars_causl}
pars <- list(A = list(beta=0),
             Z = list(beta=c(-0.2,0.4), phi=1),
             B = list(beta=c(-0.4,0.1,0.2,0.05)),
             Y = list(beta=c(-0.4,0.2,0.3,0.1), phi=0.5),
             cop = list(beta=0.5)
             )
```

```{r sim_causl}
n <- 1e3
dat <- rfrugalParam(n, formulas = forms, family = fam, pars = pars)
head(dat)
```


## Two confounding variables

```{r libs, message=FALSE}
# library(data.table)
library(DiagrammeR)
library(AIPW)
library(SuperLearner)
```

We start by selecting variables for our model. Represented by the DAG below, the following variables are in the model.

* $X$ : a binary treatment indicator
* $Y$ : a binary outcome variable
* $Z_1$ : a continuous covariate
* $Z_2$ : a binary covariate


```{r,echo=FALSE,fig.width=2.5, fig.height=2.5}
grViz("
	digraph causal {
	splines=false
	
	  # Nodes
	  node [shape = circle,
	  fixedsize = true]
	  A [label = 'X']
	  R [label = 'Z1,Z2']
	  F [label = 'Y']
	  
	  # Edges
	  edge [color = black,
	        arrowhead = vee]
	  rankdir = LR
	  A->F
	  R->A
	  R->F

	  # Graph
	  graph [overlap = true, fontsize = 10]
	}")

```

Theorem 3.1 in the paper states that we can parameterize the model using parametric models of the three pieces:

$p(x,z_1,z_2)$ , $p(y\mid do(x))$, and $\phi^*_{YZ_1Z_2\mid X}(y,z_1,z_2\mid x)$.

We therefore parameterize the distributions as below:

* $Z_1 \sim \mathcal{N}(0,1)$
* $Z_2 \sim \text{Bernoulli}(0.5)$
* $X \sim \text{Bernoulli}(\text{expit}( Z_1+ Z_2))$
* $Y\mid do(X) \sim \text{Bernoulli}(\text{expit}(X))$.

As outlined in Example R6 in the paper, we use a latent Gaussian copula model to specify the dependence between $Z_1$, $Z_2$ and $Y$, which is a combination of continuous and discrete variables. Speficially,

* $\rho_{Z_1Z_2} = 2 \text{expit}(0.1) -1$
* $\rho_{Z_1Y} = 2 \text{expit}(0.2) -1$
* $\rho_{Z_1Y} = 2 \text{expit}(0.3) -1$

We can use the `rfrugalParam()` function to simulate data from the above causal model.

```{r}

n <- 1000
# specify families of variables

family <- list(c(1,5), c(5), 5,1)

# formula
forms  <- list(c(Z1 ~ 1,Z2 ~ 1),
               list(X ~ Z1 + Z2),
               Y ~ X,
               ~ 1)

# specify parameters
pars <- list(Z1 = list(beta = 0,phi = 1),
             Z2 = list(beta = 0),
             X = list(beta = c(0,1,1)),
             Y = list(beta = c(0,1)),
             cop = list(beta = matrix(c(0.1,0.2,0.3),nrow = 1)))

dat <- rfrugalParam(n = n, formulas = forms, family = family, pars = pars)

head(dat)
```

## Check the simulated data

We use the `AIPW` package to verify the simulated causal effect. The results indicate that the causal effects were simulated as intended.

```{r}
# true values
or_true <- exp(1)
rd_true <- expit(1) - expit(0)
rr_true <- expit(1) / expit(0)
c(or_true = or_true,
  rd_true = rd_true,
  rr_true = rr_true)


aipw_est <- AIPW$new(Y = dat$Y,
                   A = dat$X,
                   W = dat[,c("Z1","Z2")], 
                   Q.SL.library = c("SL.mean","SL.speedglm"),
                   g.SL.library = c("SL.mean","SL.speedglm"),
                   k_split = 3,
                   verbose = FALSE)$fit()$summary(g.bound = 0.025)
# Odds Ratio
aipw_est$result

```


## Fitting

We can use the `fitCausal` function to fit a causal model on data.

```{r mle, cache=TRUE}
# start <- Sys.time()
causal_fit <- fitCausal(dat = dat, formulas = unlist(forms[-2]), family = unlist(family[-2]), sandwich = T)
# print(paste0("Fitting time: ",Sys.time() -start))
causal_fit
```


