---
title: "Simulating Data from Causal Models using the causl R Package"
output:
  html_document:
    toc: true
    toc_float: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `causl` R package provides a set of tools for simulating data from, and fitting causal models. In this tutorial, we will demonstrate how to use the `causl` package to simulate data.

# Install and load the `causl` Package

Before we can use the `causl` package, we need to install it. We can do this using the following code:

```{r install-causl, message=FALSE}
# devtools::install_github("rje42/causl")
library(causl)

```

```{r, echo = F, message=FALSE}
library(data.table)
library(DiagrammeR)
library(AIPW)
library(SuperLearner)

```

# Simulate Data

We start by selecting variables for our model. Represented by the DAG below, the following variables are in the model.

* $X$ : a binary treatment indicator
* $Y$ : a binary outcome variable
* $Z_1$ : a continuous covariate
* $Z_2$ : a binary covariate


```{r,echo=FALSE,fig.width=2.5, fig.height=2.5}
grViz("
	digraph causal {
	splines=false
	
	  # Nodes
	  node [shape = circle,
	  fixedsize = true]
	  A [label = 'X']
	  R [label = 'Z1,Z2']
	  F [label = 'Y']
	  
	  # Edges
	  edge [color = black,
	        arrowhead = vee]
	  rankdir = LR
	  A->F
	  R->A
	  R->F

	  # Graph
	  graph [overlap = true, fontsize = 10]
	}")

```

Theorem 3.1 in the paper states that we can parameterize the model using parametric models of the three pieces:

$p(x,z_1,z_2)$ , $p(y\mid do(x))$, and $\phi^*_{YZ_1Z_2\mid X}(y,z_1,z_2\mid x)$.

We therefore parameterize the distributions as below:

* $Z_1 \sim \mathcal{N}(0,1)$
* $Z_2 \sim \text{Bernoulli}(0.5)$
* $X \sim \text{Bernoulli}(\text{expit}( Z_1+ Z_2))$
* $Y\mid do(X) \sim \text{Bernoulli}(\text{expit}(X))$.

As outlined in Example R6 in the paper, we use a latent Gaussian copula model to specify the dependence between $Z_1$, $Z_2$ and $Y$, which is a combination of continuous and discrete variables. Speficially,

* $\rho_{Z_1Z_2} = 2 \text{expit}(0.1) -1$
* $\rho_{Z_1Y} = 2 \text{expit}(0.2) -1$
* $\rho_{Z_1Y} = 2 \text{expit}(0.3) -1$

We can use the `rfrugalParam()` function to simulate data from the above causal model.

```{r}

n <- 1000
# specify families of variables

family <- list(c(1,5), c(5), 5,1)

# formula

forms  <- list(c(Z1 ~ 1,Z2 ~ 1),
               list(X ~ Z1 + Z2),
               Y ~ X,
               ~ 1)

# specify parameters
pars <- list(Z1 = list(beta = 0,phi = 1),
             Z2 = list(beta = 0),
             X = list(beta = c(0,1,1)),
             Y = list(beta = c(0,1)),
             cop = list(beta = matrix(c(0.1,0.2,0.3),nrow = 1)))

dat <- as.data.table(causl:::rfrugalParam(n = n, formulas = forms, family = family, pars = pars))

head(dat)
```

# Check the simulated data

We use the `AIPW` package to verify the simulated causal effect. The results indicate that the causal effects were simulated as intended.

```{r}
# true values
or.true <- exp(1)
rd.true <- expit(1) - expit(0)
rr.true <- expit(1) / expit(0)
print(as.data.table(list(or.true = or.true,
                         rd.true = rd.true,
                         rr.true = rr.true)))


aipw.est <- AIPW$new(Y = dat$Y,
                   A = dat$X,
                   W = dat[,c("Z1","Z2")], 
                   Q.SL.library = c("SL.mean","SL.speedglm"),
                   g.SL.library = c("SL.mean","SL.speedglm"),
                   k_split = 3,
                   verbose = FALSE)$fit()$summary(g.bound = 0.025)
# Odds Ratio
aipw.est$result

```


# Fitting

We can use the `fitCausal` function to fit a causal model on data.

```{r}
# start <- Sys.time()
causal.fit <- fitCausal(dat = dat, formulas = unlist(forms[-2]), family = unlist(family[-2]), sandwich = T)
# print(paste0("Fitting time: ",Sys.time() -start))
causal.fit
```




